---
title: "NBA Draft Picks and Team Success - Stat 433"
author: "Sam Murray, Madison Dunn, Jake Reiner,  Huailin Tang"
date: "11/17/2021"
output: html_document
---

```{r echo = FALSE, message=FALSE, warning=FALSE}
library("RSQLite")
library(tidyverse)
library("ggfortify")
library(ggplot2)
library("caret")
library(shiny)
```
Link to github: https://github.com/madisondunn8/433-app

- Note: run shiny app by putting **runGitHub("433-app", "madisondunn8",ref="main")** into RStudio. Be sure to call library(shiny) first. 

- Original data from: https://www.kaggle.com/wyattowalsh/basketball and https://www.kaggle.com/boonpalipatana/nba-season-records-from-every-year

**For our project, we decided to investigate the relationship between draft picks and the NBA. The data used can be seen in a shiny app, titled "App.R". Our full written report is also in this repository.**

```{r message=FALSE, warning=FALSE}
con <- dbConnect(drv=RSQLite::SQLite(), dbname="basketball.sqlite")

tables <- dbListTables(con)

tables <- tables[tables != "sqlite_sequence"]

lDataFrames <- vector("list", length=length(tables))

for (i in seq(along=tables)) {
  lDataFrames[[i]] <- dbGetQuery(conn=con, statement=paste("SELECT * FROM '", tables[[i]], "'", sep=""))
}

draft_data = lDataFrames[[1]]
combine_data = lDataFrames[[2]]
season_data=lDataFrames[[3]]

wins_data = read.csv("Team_Records.csv")
colnames(wins_data)[1] = "Season"
```

```{r, inclue=FALSE}
# combine draft data with team success data merging by team and year

draft = draft_data %>% 
  mutate(teamYear = paste(teamName, ", ", yearDraft, sep=""))

wins_data = wins_data %>% 
  mutate(teamYear = paste(gsub("[^[:alnum:]]", "", word(Team,-1)), ", ", substr(Season, 1, 4), sep=""))

draft_and_wins = merge(draft, wins_data, by.x = "teamYear", by.y = "teamYear")
```


```{r, echo=FALSE}
df1 = wins_data[1,]
five_after_first = list()
#rbind(df1, df2)
index = 0
for (i in draft_and_wins[draft_and_wins$numberPickOverall==1,]$teamYear){
  team = strsplit(i, ', ')[[1]][1]
  year = as.numeric(strsplit(i, ', ')[[1]][2])
  
  df2 = wins_data[1,]
  for (j in 1:5){
    a = wins_data[wins_data$teamYear == paste(team, ", ", toString(year + j), sep=""),]
    
    df1 = rbind(df1, a)
    df2 = rbind(df2, a)
    #print(b)
    ##points(a$W.L., a$yearDraft)
    
  }
  index = index + 1
  five_after_first[[index]] = df2[2:6, ]
  df2 = ""
  
}
five_years_after_first_pick = df1 %>% distinct(teamYear, .keep_all = TRUE)
```

```{r, inclue=FALSE}
draft_scores = draft_and_wins %>% 
  filter(yearDraft>=1990) %>% 
  group_by(teamYear) %>% 
  summarize(pick_sum = sum(numberPickOverall), 
            pick_count = n(), 
            best_pick = min(numberPickOverall),
            worst_pick = max(numberPickOverall),
            draft_score = 61*pick_count - pick_sum) %>%
  arrange(desc(draft_score))
```

```{r, echo = FALSE}
head(draft_scores)
```

```{r, include=FALSE}
five_after = list()
index = 0

for (i in draft_scores$teamYear){
  team = strsplit(i, ', ')[[1]][1]
  year = as.numeric(strsplit(i, ', ')[[1]][2])
  
  draft_score = as.numeric(draft_scores[draft_scores$teamYear == i,]$draft_score)
  best_pick = as.numeric(draft_scores[draft_scores$teamYear == i,]$best_pick)
  win_pct = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year), sep=""),]$W.L.)
  win_pct_2 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 2), sep=""),]$W.L.)
  win_pct_3 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 3), sep=""),]$W.L.)
  win_pct_4 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 4), sep=""),]$W.L.)
  win_pct_5 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 5), sep=""),]$W.L.)
  win_pct_6 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 6), sep=""),]$W.L.)
  win_pct_7 = as.double(wins_data[wins_data$teamYear == paste(team, ", ", toString(year + 7), sep=""),]$W.L.)
  
  avg_win_pct_change = ((win_pct_2 - win_pct) + (win_pct_3 - win_pct_2) + (win_pct_4 - win_pct_3) + (win_pct_5 - win_pct_4) + (win_pct_6 - win_pct_5) + (win_pct_7 - win_pct_6)) / 6
  
  win_pct_change_3_years = win_pct_4 - win_pct
  win_pct_change_5_years = win_pct_6 - win_pct
  
  if (length(win_pct_3) == 1){
    index = index + 1
    five_after[[index]] = c(team, year, draft_score, best_pick, win_pct_change_3_years, win_pct_change_5_years, win_pct, win_pct_2, win_pct_3, win_pct_4, win_pct_5, win_pct_6, win_pct_7, avg_win_pct_change)
  }
}

draft_scores_records = as.data.frame(do.call(rbind, five_after))
colnames(draft_scores_records) = c("team", "year", "draft_score", "best_pick", "win_pct_change_3_years", "win_pct_change_5_years", "win_pct_current", "win_pct_2_years_after", "win_pct_3_years_after", "win_pct_4_years_after", "win_pct_5_years_after", "win_pct_6_years_after", "win_pct_7_years_after",  "avg_win_pct_change")
draft_scores_records$draft_score <- as.integer(as.character(draft_scores_records$draft_score))
draft_scores_records$win_pct_current <- as.double(as.character(draft_scores_records$win_pct_current))
draft_scores_records$win_pct_2_years_after <- as.double(as.character(draft_scores_records$win_pct_2_years_after))
draft_scores_records$win_pct_3_years_after <- as.double(as.character(draft_scores_records$win_pct_3_years_after))
draft_scores_records$win_pct_4_years_after <- as.double(as.character(draft_scores_records$win_pct_4_years_after))
draft_scores_records$win_pct_5_years_after <- as.double(as.character(draft_scores_records$win_pct_5_years_after))
draft_scores_records$win_pct_6_years_after <- as.double(as.character(draft_scores_records$win_pct_6_years_after))
draft_scores_records$win_pct_7_years_after <- as.double(as.character(draft_scores_records$win_pct_7_years_after))
draft_scores_records$win_pct_change_3_years <- as.double(as.character(draft_scores_records$win_pct_change_3_years))
draft_scores_records$win_pct_change_5_years <- as.double(as.character(draft_scores_records$win_pct_change_5_years))
draft_scores_records$best_pick <- as.integer(as.character(draft_scores_records$best_pick))
draft_scores_records$year <- as.integer(as.character(draft_scores_records$year))
draft_scores_records$avg_win_pct_change <- as.double(as.character(draft_scores_records$avg_win_pct_change))
#draft_scores_records[, 2:6] <- sapply(draft_scores_records[, 3:6], as.double)
draft_scores_records = draft_scores_records %>% filter(win_pct_2_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(win_pct_3_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(win_pct_4_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(win_pct_5_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(win_pct_6_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(win_pct_7_years_after <= 1.00)
draft_scores_records = draft_scores_records %>% filter(avg_win_pct_change <= 200)
```

```{r, echo=FALSE}
###############
# Shiny app
###############
#runGitHub("433-app","madisondunn8", ref = "main")
```

```{r, echo=FALSE}
draft_scores_records %>% 
  filter(draft_score >= 80) %>%
  ggplot(aes(x = draft_score, y = avg_win_pct_change)) + 
  geom_point()

draft_scores_records %>% 
  filter(best_pick <= 3) %>%
  ggplot(aes(x = draft_score, y = win_pct_5_years_after)) + 
  geom_point()

draft_scores_records %>% 
  #filter(best_pick <= 30) %>%
  ggplot(aes(x = draft_score- best_pick, y = win_pct_change_3_years )) + 
  geom_point()
```




***<h1>Test Linear Model using draft score to predict change in team win percentage from the draft year to 5 years after the draft</h1>***



```{r}
# all teams
all <- lm(win_pct_change_3_years ~ draft_score + best_pick, data = draft_scores_records)
plot(x=predict(all), y=draft_scores_records$win_pct_change_3_years)
summary(all)
```

- **Best pick is statistically significant here but draft score is not.**

```{r}
# teams below 500

below_500 = draft_scores_records %>% 
  filter(win_pct_current < 0.5)

bad <- lm(win_pct_change_5_years ~ draft_score, data = below_500)
summary(bad)
plot(x=predict(bad), y=below_500$win_pct_change_3_years)
```

```{r}
# teams with a top 10 pick
top_10 = draft_scores_records %>% 
  filter(best_pick <= 10)

top <- lm(win_pct_change_5_years ~ draft_score, data = top_10)
summary(top)
plot(x=predict(top), y=top_10$win_pct_change_3_years)
```


Figure One shows and NBA teams draft score in a given year versus the teams success 5 years after the scored draft. Draft scores are calculated by taking 61 minus the pick number the team has (60 total picks in draft), summed for all picks for a team. This figure is significant because this is the analysis we are trying to achieve in our project, if draft capital improves NBA teams. Although not much of a correlation can be seen in the plot, the relationship between these variables is what we are interested in our project. 





```{r}
log_model = lm(win_pct_change_3_years ~ log(draft_score)+best_pick, data = draft_scores_records)

summary(log_model)

ggplot(draft_scores_records, aes(draft_score, win_pct_change_3_years)) +
  geom_point() +
  stat_smooth(method=lm,formula=y~log(x))
```


- Draft score is statistically significant at a .10 level after taking a log transformation of the variable. 

